Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-19 _tex/manuscript.tex
Built site for gh-pages

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

@@

To be able to create models for lymphatic tumor progression that
includes all relevant LNLs including the contralateral side, we have
collected a detailed dataset of 833 patients with newly diagnosed
oropharyngeal squamous cell carcinomas. It reports the lymph node
involvement per LNL of every patient individually in tabular form, in
addition to other primary tumor and patient characteristics such as
T-category, subside, and lateralization of the primary tumor, and HPV
p16 status. Their patient records have been collected at four different
institutions and a brief overview over some of their patients'
characteristics are shown in table~\ref{tbl-data-overview}. The data
from the Inselspital Bern and the Centre Léon Bérard only consists of
patients who received a neck dissection. Since this treatment is more
commonly chosen for early T-category patients, they also make up a
larger portion of the respective dataset. The majority of patients from
the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
was treated with definitive radiotherapy.

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

@@

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript-preview.html}{Article
Notebook}}

@@

\includegraphics{manuscript_files/mediabag/figures/fig-data-strat.pdf}

@@

Similarly, ipsilateral involvement correlates with contralateral
metastasis. The tumor of a patients with many metastases in ipsilateral
LNLs was probably able to spread for longer (or faster) compared to a
tumor in a patient with no nodal disease. This, too, may therefore be
considered a surrogate for the duration of the disease. In rare cases,
bulky nodal disease ipsilaterally may also redirect lymph fluids to the
contralateral side.

Lastly, the right panel in figure~\ref{fig-data-strat} shows that
patients with a tumor crossing the mid-sagittal line show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intutitive sense, because the lymphatic
system in the head and neck region is typically symmetric and thus no
major vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multi-centric_2023}. Although these publications do not
include the most recent dataset addition from Vall d'Hebron Barcelona
Hospital.

@@

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/small-graph.pdf}

@@

\caption{\label{fig-small-graph}Directed acyclic graph (DAG)
representing the abstract lymphatic network in the head and neck region.
Blue nodes are the LNLs' hidden random variables, the red node
represents the tumor, and the orange square nodes depict the binary
observed variables. Red and blue arcs symbolize the probability of
lymphatic spread along that edge during one time-step. The orange arcs
represent the sensitivity and specificity of the observational modality
(e.g.~CT, MRI, pathology, \ldots).}

@@

Our first model to predict the lymphatic progression of HNSCC was
introduced using Bayesian networks \citep{pouymayou_bayesian_2019}. We
subsequently extended this work to a hidden Markov model (HMM)
\citep{ludwig_hidden_2021} to allow an intuitive inclusion of T-category
into the predictions. We will briefly summarize this HMM's formalism
before building on it to include the contralateral spread in
section~\ref{sec-ext-to-contra}.

Here, \(V\) is the number of LNLs the model considers. The values a
LNL's hidden binary RV may take on are \(X_v[t] = 0\) (\texttt{False}),
meaning the LNL \(v\) is healthy or free of metastatic disease, or
\(X_v[t] = 1\) (\texttt{True}), corresponding to some for of tumor
presence (i.e., occult or clinical).

Since the state vector \(\mathbf{X}[t]\) is \(V\)-dimensional and
binary, there are \(2^V\) distinct possible lymphatic involvement
patterns, which we enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A starting state \(\mathbf{X}[t=0]\) at time \(t=0\) just before the
  patient's tumor formed. In our case, this is always the state where
  all LNLs are still healthy \(\boldsymbol{\xi}_0\).
\item
  The \emph{transition matrix}
  \begin{equation}\phantomsection\label{eq-trans-matrix}{
  \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where the value at row \(i\) and column \(j\)
  represents the probability to transition from state
  \(\boldsymbol{\xi}_i\) to \(\boldsymbol{\xi}_j\) during the time-step
  from \(t\) to \(t+1\). Note that we prohibit self-healing, meaning
  that during a transition, no LNL may change their state from
  \(X_v[t]=1\) to \(X_v[t+1]=0\). This effectively masks large parts of
  the transition matrix to be zero.
\item
  Lastly, the \emph{observation matrix}
  \begin{equation}\phantomsection\label{eq-obs-matrix}{
  \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where in row \(i\) and at column \(j\) we find the
  probability to \emph{observe} a lymphatic involvement pattern
  \(\mathbf{Z} = \boldsymbol{\zeta}_j\), given that the true (but
  hidden) state of involvement at the time of diagnosis \(t_D\) is
  \(\mathbf{X}[t_D] = \boldsymbol{\xi}_i\).
\end{enumerate}

Note that the transition matrix \(\mathbf{A}\) is parametrized using the
spread probabilities of a directed acyclic graph (DAG) that we define as
the underlying mechanistic representation of the lymphatic network. An
example of such a DAG is shown in figure~\ref{fig-small-graph}.

Using the introduced quantities, we can evolve the distribution of all
possible hidden states from \(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\)
step by step, by successively multiplying this vector with the
transition matrix \(\mathbf{A}\). At the time of the diagnosis \(t_D\),
we multiply the result with the observation matrix \(\mathbf{B}\). We
may then look up the likelihood of a patient presenting with the
diagnosis \(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of
the final result.

However, the remaining issue is that the value of \(t_D\) is unknown,
i.e.~over how many time-steps the HMM should be evolved. We solve this
problem by marginalizing over the time of diagnosis. Different
distributions over the diagnosis times can then be choosen based on
T-category. For instance, the mean of the time-prior to marginalize over
the diagnosis time for early T-category patients
\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
times than the one for advanced T-category patients
\(P\left( t_D \mid \text{early} \right)\). This gives us for example

@@

For later use, we define at this point the matrices of the dsitributions
over all hidden states, given all time-steps \(\boldsymbol{\Lambda}\):

Where the \(k\)-th row in this matrix corresponds to the distribution
over hidden states after \(t=k-1\) time-steps.

@@

The single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-small-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, the
parameter space of the unilateral model is:

@@

A naive approach to model the contralateral lymphatic spread would be to
simply employ two independent unilateral models as introduced in
section~\ref{sec-unilateral}. During training, one could even enforce
some shared parameters between these two models, like the
parameterization of the distributions over diagnose times or the spread
among the LNLs. Additionally, we could think of an approach to
incorporate the primary tumor's mid-sagittal extension as a risk factor.

However, this approach lacks a way to describe the correlation between
ipsi- and contralateral involvement. This is displayed in
table~\ref{tbl-data-strat} and shows how often the contralateral LNLs I,
II, III, and IV were involved, given all possible combinations of
midline extension, T-category, and ipsilateral LNL III involvement.
Unsurprisingly, the prevalence for contralateral involvement is
consistently higher when the tumor extends over the mid-sagittal line or
is of later T-category. But it is also more frequent when the
ipsilateral side shows more severe involvement, which is here shown via
the surrogate LNL III.

Thus, we attempt to extend the formalism in section~\ref{sec-unilateral}
in such a way that the model's ipsi- and contralateral side evolve
synchronously. To achieve that, we start by writing down the posterior
distribution of involvement an analogy to
equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

@@

The likelihood of the diagnoses given a hidden state simply factorise:
\(P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)\).
And the two factors are contained in the observation matrices
\(\mathbf{B}^\text{i}\) and \(\mathbf{B}^\text{c}\).

The term representing the model's prior probability of hidden
involvement does not factorize. However, if we assume that lymphatic
spread typically does not cross the mid-sagittal line, we can write it
as a factorising sum:

@@

This assumption makes intuitive sense: The two sides of the lymphatic
network in a typical patient are approximately mirror images of each
other. Thus, no major vessels cross the mid-sagittal line. There may,
however, be diffusion of lymph fluid accross this line or bulky
involvement that redirects lymphatic drainage significantly.

Using this assumption along with equation~\ref{eq-lambda-matrix}, we can
write the above distribution algebraically as a product:

@@

In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
disjoint set of parameters. I.e., the ipsi- and contralateral spread
rates are entirely different. However, using two sensible assumptions,
we can reduce the parameter space by sharing some parameters between the
sides:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We assume the spread \emph{among} the LNLs to be same on both sides.
  It is reasonable to assume the lymphatic system is symmetric. Thus,
  the spread rates from one LNL to the other should be symmetric, too.
  Formally, this means\\
  \begin{equation}\phantomsection\label{eq-symmetries}{
  \begin{aligned}
  b_v^\text{c} &\neq b_v^\text{i} \\
  t_{rv}^\text{c} &= t_{rv}^\text{i}
  \end{aligned}
  }\end{equation} for all \(v \leq V\) and
  \(r \in \operatorname{pa}(v)\).
\item
  The tumor's spread to the contralateral side in case of an extension
  over the midline is larger than if it was clearly lateralized, but
  smaller than its spread to the ipsilateral side. This assumption stems
  from a simple thought experiment: Consider moving the tumor from a
  clearly lateralized position accross the mid-sagittal plane to the
  same position, but on the contralateral side. In the beginning we
  would have \(b_v^\text{c} < b_v^\text{i}\), while in the end, the
  situation is reversed. If a tumor extends over the mid-sagittal line,
  its contralateral spread rate can be expected to be in between these
  two extremes. We encode this in a \emph{mixing parameter} \(\alpha\)
  that captures a ``degree of asymmetry'':\\
  \begin{equation}\phantomsection\label{eq-mixing}{
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  }\end{equation} This means the model now uses three different sets of
  parameters to describe the spread from the tumor to the LNLs:
  \(b^\text{i}_v\) for the spread to the ipsilateral LNLs,
  \(b_v^{\text{c},\epsilon=\texttt{False}}\) for the spread to the
  contralateral LNLs as long as the tumor is clearly lateralized, and
  finally \(b_v^{\text{c},\epsilon=\texttt{True}}\) when it crosses the
  midline. Note, however, that these three sets of spread rates only
  account for \(2 \cdot 2^V + 1\) parameters, since they are coupled via
  the mixing parameter \(\alpha\).
\end{enumerate}

@@

which is less than double in size, compared to the unilateral model.
From these parameters, we now build three different transition matrices:
The unchanged \(\mathbf{A}_\text{i}\) for the ipsilateral side, one
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the contralateral
side that covers the progression as long as the tumor is lateralized,
and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in case of a tumor
that has crossed the mid-sagittal line.

@@

To account for the increased prevalence of involvement on the
contralateral side when the tumor is not clearly lateralized anymore, we
also model the tumor's extension over the mid-sagittal line as a binary
random variable. It starts lateralized and at every time-step there is a
finite probability \(p_\epsilon\) that the tumor grows over the symmetry
plane of the patient.

The overall probabilities to find a patient with a clearly lateralized
tumor or one that extends over the mid-sagittal line after \(t\)
time-steps are then respectively given by

>>>>

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
Notebook}}

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
Notebook}}

@@

To be able to create models for lymphatic tumor progression that include
all relevant LNLs including the contralateral side, we have collected a
detailed dataset of 833 patients with newly diagnosed oropharyngeal
squamous cell carcinomas
\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
lymph node involvement per LNL of every patient individually in tabular
form, in addition to other primary tumor and patient characteristics
such as T-category, subside, and lateralization of the primary tumor,
and HPV p16 status. Their patient records have been collected at four
different institutions and a brief overview over some of their patients'
characteristics are shown in table~\ref{tbl-data-overview}. The data
from the Inselspital Bern and the Centre Léon Bérard only consist of
patients who received a neck dissection. Since this treatment is more
commonly chosen for early T-category patients, they also make up a
larger portion of the respective dataset. The majority of patients from
the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
was treated with definitive radiotherapy. For a subset of 83 patients in
the CLB dataset, the lateralization of the primary tumor was not
reported.

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
Notebook}}

@@

\textsubscript{Source:
\href{https://rmnldwg.github.io/bilateral-paper/manuscript.qmd.html}{Article
Notebook}}

\subsection{Consensus on Involvement
Status}\label{consensus-on-involvement-status}

Pathological involvement is thus available only for a subset of the
patients and only for the dissected levels. For the remaining patients,
only clinical involvement is available. For the analysis performed in
this paper, all available diagnostic information has been combined into
a consensus decision representing the most likely state of involvement
for each patient and LNL, using literature values for sensitivity and
specificity of different modalities
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.
These values are tabulated in table~\ref{tbl-spec-sens}.

This consensus was formed as follows: Suppose the involvement status
\(X_v\) of LNL \(v\) was assessed using different diagnostic modalities
\(\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}\),
each characterized by their own pair of sensitivity and specificity
values \(s_N^{\mathcal{o}}\) and \(s_P^{\mathcal{o}}\), with
\(\mathcal{o} \in \mathcal{O}\). Then we have \(|\mathcal{O}|\)
observations \(z_v^{\mathcal{o}} \in \left[ 0, 1 \right]\), where 0
stands for ``healthy'' and 1 for ``involved''. We can then compute the
most likely true involvement \(X_v\) using the likelihood function

\[
\begin{aligned}
\ell \left( X_v \mid \{ z_v^{\mathcal{o}} \}_{\mathcal{o} \in \mathcal{O}} \right) = \prod_{\mathcal{o} \in \mathcal{O}}
\left( 1 - X_v \right) \cdot &\left[ z_v^{\mathcal{o}} \cdot \left( 1 - s_P^{\mathcal{o}} \right) + \left( 1 - z_v^{\mathcal{o}} \right) \cdot s_P^{\mathcal{o}} \right] \\
+ X_v \cdot &\left[ z_v^{\mathcal{o}} \cdot s_N^{\mathcal{o}} + \left( 1 - z_v^{\mathcal{o}} \right) \cdot (1 - s_N^{\mathcal{o}}) \right]
\end{aligned}
\]

We now assume the true state \(X_v\) to take on the value 1 if
\(\ell \left( X_v = 1 \mid \ldots \right) > \ell \left( X_v = 0 \mid \ldots \right)\)
and 0 otherwise. For example, if we have \(z_\text{II}^\text{CT} = 0\)
and \(z_\text{II}^\text{MRI} = 1\) we would compute the following
likelihoods:

\[
\begin{aligned}
\ell \left( X_\text{II} = 1 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= \left( 1 - s_N^\text{CT} \right) \cdot s_N^\text{MRI} = 15.39\% \\
\ell \left( X_\text{II} = 0 \mid z_\text{II}^\text{CT} = 0, z_\text{II}^\text{MRI} = 1 \right) &= s_P^\text{CT} \cdot \left(1 - s_N^\text{MRI}\right) = 14.44\%
\end{aligned}
\]

In this example, we would thus assume the true state to be involved
(\(X_\text{II} = 1\)).

This method of computing a consensus also ensures that the pathology
reports always override any conflicting clinical diagnosis, due to
pathology's high sensitivity and specificity.

\begin{longtable}[]{@{}lrr@{}}
\caption{Specificity and sensitivity values from the literature
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.}\label{tbl-spec-sens}\tabularnewline
\toprule\noalign{}
Modality & Specificity & Sensitivity \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Modality & Specificity & Sensitivity \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CT & 76\% & 81\% \\
PET & 86\% & 79\% \\
MRI & 63\% & 81\% \\
FNA & 98\% & 80\% \\
pathology & 100\% & 100\% \\
\end{longtable}

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multi-centric_2023}. The Data-in-Brief publications
contain a detailed description of the datasets and the data format.
However, these publications do not include the most recent dataset
addition from Vall d'Hebron Barcelona Hospital.

@@

\includegraphics{manuscript_files/mediabag/figures/fig_data_strat.pdf}

@@

Similarly, ipsilateral involvement correlates with contralateral
metastasis. The tumor of a patients with many metastases in ipsilateral
LNLs was probably able to spread for longer (or faster) compared to a
tumor in a patient with no nodal disease. This, too, may therefore be
considered a surrogate for the duration of the disease. In addition, it
has been hypothesized that bulky nodal disease ipsilaterally may also
redirect lymph fluids to the contralateral side.

Lastly, the right panel in figure~\ref{fig-data-strat} shows that
patients with a tumor crossing the mid-sagittal line show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intuitive sense, because the lymphatic
system in the head and neck region is typically symmetric and thus no
major vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 48\% of advanced T-category tumors but
only in 7\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, figure~\ref{fig-data-strat-uncorr} indicates that
advanced T-category and severe ipsilateral involvement do represent
additional risk factors . The figure considers only patients with
lateralized tumors not extending over the midline. For early T-category
and healthy ipsilateral levels I-V, only 1\% (1 out of 84 patients)
shows involvement of contralateral level II. This increases to 5\% (9
out of 165 patients) if ipsilateral level II is involved, to 15\% (9 out
of 60 patients) if ipsilateral levels II and III are involved, and to
23\% (6 out of 26 patients) for advanced T-category with ipsilateral
levels II and III involved.

\includegraphics{manuscript_files/mediabag/figures/fig_data_strat_uncorr.pdf}

\caption{\label{fig-data-strat-uncorr}Contralateral involvement
prevalence by LNL for lateralized tumors. Shown are four different
scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2)
early T-category with ipsilaterally LNL II involved (blue), 3) late
T-category with ipsilateral LNL II metastatic (orange), and 4) early
T-category with ipsilateral both levels II and III involved (red). This
illustrates that even when controlling for the tumor's lateralization,
T-category and ipsilaeral involvement are predictive of contralateral
metastases.}

@@

@@

This paper builds on the previously developed unilateral model for
ipsilateral lymph node involvement presented in {[}cite 2024 graph
extension paper{]}. In this section we provide a brief recap of the
unilateral model to introduce the notation needed to extend the
framework to a bilateral model describing both ipsilateral and
contralateral lymph node involvement in section~\ref{sec-ext-to-contra}.
For further details in the ipsilateral model, the reader is referred to
the earlier publications \citep[cite 2024 graph extension
paper]{ludwig_hidden_2021}

Here, \(V\) is the number of LNLs the model considers. The values a
LNL's hidden binary random variables may take on are \(X_v[t] = 0\)
(\texttt{False}), meaning the LNL \(v\) is healthy or free of metastatic
disease, or \(X_v[t] = 1\) (\texttt{True}), corresponding to some form
of tumor presence (i.e., occult or clinically detected). Since the state
vector \(\mathbf{X}[t]\) is \(V\)-dimensional and binary, there are
\(2^V\) distinct possible lymphatic involvement patterns, which we
enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
In addition, each LNL is associated with an observed binary random
variable \(Z_v\) that describes the clinical involvement of a LNL based
on imaging: \(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is
healthy based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True})
indicates that suspicious lymph nodes were detected that are deemed
metastatic.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A starting state \(\mathbf{X}[t=0]\) at time \(t=0\) just before the
  patient's tumor formed. In our case, this is always the state where
  all LNLs are still healthy \(\boldsymbol{\xi}_0\).
\item
  The \emph{transition matrix}
  \begin{equation}\phantomsection\label{eq-trans-matrix}{
  \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where the value at row \(i\) and column \(j\)
  represents the probability to transition from state
  \(\boldsymbol{\xi}_i\) to \(\boldsymbol{\xi}_j\) during the time-step
  from \(t\) to \(t+1\). Note that we prohibit self-healing, meaning
  that during a transition, no LNL may change their state from
  \(X_v[t]=1\) to \(X_v[t+1]=0\). Consequently, many elements of the
  transition matrix are zero.
\item
  Lastly, the \emph{observation matrix}
  \begin{equation}\phantomsection\label{eq-obs-matrix}{
  \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where in row \(i\) and at column \(j\) we find the
  probability to \emph{observe} a lymphatic involvement pattern
  \(\mathbf{Z} = \boldsymbol{\zeta}_j\), given that the true (but
  hidden) state of involvement at the time of diagnosis \(t_D\) is
  \(\mathbf{X}[t_D] = \boldsymbol{\xi}_i\).
\end{enumerate}

The transition matrix \(\mathbf{A}\) is parametrized using a directed
acyclic graph (DAG) as an abstract representation of the underlying
lymphatic network. Directed arcs from the primary tumor to a LNL are
associated with a probability \(b_v\) for direct spread to LNL \(v\)
during one time step. Directed arcs from a LNL \(v\) to a LNL \(r\) are
associated with a probability \(t_{vr}\) for the tumor to progress from
one LNL to the next. In this paper, we build on the DAG shown in
figure~\ref{fig-full-graph} which was obtained by maximizing the model
evidence as described in {[}cite 2024 graph extension paper{]}.

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/full-graph.pdf}

@@

\caption{\label{fig-full-graph}Directed acyclic graph (DAG) representing
the abstract lymphatic network in the head and neck region. Blue nodes
are the LNLs' hidden random variables, the red node represents the
tumor, and the orange square nodes depict the binary observed variables.
Red and blue arcs symbolize the probability of lymphatic spread along
that edge during one time-step. The orange arcs represent the
sensitivity and specificity of the observational modality (e.g.~CT, MRI,
pathology, \ldots).}

@@

Using the introduced quantities, we can evolve the distribution of all
possible hidden states from \(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\)
step by step, by successively multiplying this vector with the
transition matrix \(\mathbf{A}\). For later use, we define at this point
the matrices of the dsitributions over all hidden states, given all
time-steps \(\boldsymbol{\Lambda}\):

Where the \(k\)-th row in this matrix corresponds to the probability
distribution over hidden states after \(t=k-1\) time-steps.

At the time of the diagnosis \(t_D\), we multiply the result with the
observation matrix \(\mathbf{B}\). We may then look up the likelihood of
a patient presenting with the diagnosis
\(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of the final
result. However, the remaining issue is that the value of \(t_D\) is
unknown, i.e.~over how many time-steps the HMM should be evolved. We
solve this problem by marginalizing over the time of diagnosis.
Different distributions over the diagnosis times can then be choosen
based on T-category. For instance, the mean of the time-prior to
marginalize over the diagnosis time for early T-category patients
\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
times than the one for advanced T-category patients
\(P\left( t_D \mid \text{early} \right)\). This gives us for example

@@

@@

The single-patient likelihood \(\ell\) in
equation~\ref{eq-single-patient-llh} depends on the spread parameters
shown in figure~\ref{fig-full-graph} via the transition matrix
\(\mathbf{A}\) and on the binomial parameters \(p_{\text{T}x}\) via
time-priors. In this work, we will only differentiate between ``early''
(T1 \& T2) and ``advanced'' (T3 \& T4) T-categories. Therefore, the
parameter space of the unilateral model is:

@@

A naive approach to model the contralateral lymphatic spread would be to
simply employ two independent unilateral models as introduced in
section~\ref{sec-unilateral}. During training, one could enforce that
some parameters are shared between these two models, e.g.~the
parameterization of the distributions over diagnose times or the spread
among the LNLs (\$t\_\{vr\}). However, this approach lacks a way to
describe the correlation between ipsi- and contralateral involvement
discussed in section \{\#sec-data-strat\}. Two independent unilateral
models would not describe the observation that contralateral involvement
becomes more likely with more severe ipsilateral involvement.

Thus, we extend the formalism in section~\ref{sec-unilateral} in such a
way that the model's ipsi- and contralateral side evolve synchronously
over time. To achieve that, we start by writing down the posterior
distribution of involvement an analogy to
equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

@@

The probability of the diagnoses given a hidden state factorises:
\(P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)\),
and the two factors are described through observation matrices
\(\mathbf{B}^\text{i}\) and \(\mathbf{B}^\text{c}\).

The term representing the model's prior probability of hidden
involvement does not factorize. However, we assume that there is no
direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No
major lymph vessels cross the mid-sagittal plane. In the graphical
model, this means that we assume no directed arcs between ipsilateral
and contralateral LNLs, i.e.~tumor spread to the contralateral side is
due to the primary tumor alone. We can thus write the joint probability
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) as a
factorising sum:

@@

This assumption makes intuitive sense: The state of the ipsilateral and
contralateral sides of the lymphatic network evolves independently over
time as there are no major lymph vessels crossing the midline. However,
both sides are coupled via time. Qualitatively speaking, according to
the model in \{\#eq-bilateral-marginal\}, a joint state with severe
contralateral involvement and limited ipsilateral involvement would be
unlikely, because severe contralateral involvement becomes likely only
at later time steps when limited ipsilateral involvement becomes
unlikely.

Using \{\#eq-bilateral-marginal\} along with
equation~\ref{eq-lambda-matrix}, we can write the above distribution
algebraically as a product:

@@

In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
disjoint set of parameters. I.e., the ipsi- and contralateral spread
rates are entirely different. However, using three sensible assumptions,
we can reduce the parameter space by sharing some parameters between the
sides:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We assume that both ipsilateral and contralateral spread is described
  through the same graph shown in \{\#fig-full-graph\}.
\item
  We assume the spread \emph{among} the LNLs to be same on both sides.
  It is reasonable to assume the lymphatic system is symmetric. Thus,
  the spread rates from one LNL to the other should be symmetric, too.
  Formally, this means\\
  \begin{equation}\phantomsection\label{eq-symmetries}{
  \begin{aligned}
  b_v^\text{c} &\neq b_v^\text{i} \\
  t_{rv}^\text{c} &= t_{rv}^\text{i}
  \end{aligned}
  }\end{equation} for all \(v \leq V\) and
  \(r \in \operatorname{pa}(v)\).
\item
  The tumor's spread to the contralateral side in case of an extension
  over the midline is larger than if it was clearly lateralized, but
  smaller than its spread to the ipsilateral side. This assumption stems
  from a simple thought experiment: Consider moving the tumor from a
  clearly lateralized position accross the mid-sagittal plane to the
  same position, but on the contralateral side. In the beginning we
  would have \(b_v^\text{c} < b_v^\text{i}\), while in the end, the
  situation is reversed. If a tumor extends over the mid-sagittal line,
  its contralateral spread rate can be expected to be in between these
  two extremes. We encode this in a \emph{mixing parameter}
  \(\alpha \in [0,1]\) that captures a ``degree of asymmetry'':\\
  \begin{equation}\phantomsection\label{eq-mixing}{
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  }\end{equation} This means the model now uses three different sets of
  parameters to describe the spread from the tumor to the LNLs:
  \(b^\text{i}_v\) for the spread to the ipsilateral LNLs,
  \(b_v^{\text{c},\epsilon=\texttt{False}}\) for the spread to the
  contralateral LNLs as long as the tumor is clearly lateralized, and
  finally \(b_v^{\text{c},\epsilon=\texttt{True}}\) when it crosses the
  midline. Note, however, that these three sets of spread rates only
  account for \(2 \cdot 2^V + 1\) parameters, since they are coupled via
  the mixing parameter \(\alpha\).
\end{enumerate}

@@

which is less than double in size, compared to the unilateral model.
From these parameters, we now build three different transition matrices:
The unchanged matrix \(\mathbf{A}_\text{i}\) for the ipsilateral side,
one \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the
contralateral side that covers the progression as long as the tumor is
lateralized, and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in
case of a tumor that has crossed the mid-sagittal line.

@@

It can be assumed that most tumors that cross the midline at the time of
diagnose have started as lateralized tumors and grew over the midline at
a later time step. Hence, the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) only applies for a
subset of the time steps. Therefore, we also model the tumor's extension
over the mid-sagittal line as a binary random variable. A tumor starts
lateralized and at every time-step there is a finite probability
\(p_\epsilon\) that the tumor grows over the midsaggital plane. The
overall probabilities to find a patient with a clearly lateralized tumor
or one that extends over the mid-sagittal line after \(t\) time-steps
are then respectively given by
